---
title: "Hartford Strategic Vision 2020"
author: John
date: '2021-01-22'
output: html_notebook
---

Victorious Hartford
===================

## A Strategic Vision

```{r setup, cache = FALSE, echo = FALSE, include = FALSE}
options(warn = -1)
working_dir <- getwd()
print(working_dir)
if (!require("knitr")) {
  install.packages("knitr")
  library("knitr")
}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)

if (!require(akima) || !require(magick)  || !require(png)  || !require(reshape2) || !require(viridis)) {
    install.packages(c("akima", "ggplot2", "magick", "png", "reshape2", "viridis"))
    devtools::install_github("tylermorganwall/rayimage@7a9a138e10e19119c88e960f9cfb191d1fdae002")
    devtools::install_github("tylermorganwall/terrainmeshr@e112055e47033508cc45c8246b8dc0a0e94920f7")
    devtools::install_github("tylermorganwall/rayshader@d0c9bd94be95c44eff6e7d8da5eadff070dc11db")
    devtools::install_github("MilesMcBain/slippymath")
    devtools::install_github("neilcharles/geoviz")
}
```

```{r geoviz_test, cache = FALSE, dev.args = list(bg = 'transparent'), echo = FALSE}
options(warn = -1)
suppressMessages(library(geoviz))
suppressMessages(library(rayshader))
suppressMessages(library(raster))
suppressMessages(library(slippymath))

# Load an example IGC (GPS track log) file
igc <- example_igc()

# Load a small example elevation raster showing a piece of the English Lake district
DEM <- example_raster()

# import desert heightmap demo
if (file.exists("data/57-Elevation_DEMHF0p7m2016_N5145E281.img")) {
    localdem <- raster::raster("data/57-Elevation_DEMHF0p7m2016_N5145E281.img")
} else {
    localdem <- file.choose()
}

DEM <- raster(localdem)

sunangle <- 270

zscale <- 25

# Get a Stamen map that will cover our DEM
stamen_overlay <- slippy_overlay(DEM, image_source = "stamen", image_type = "watercolor", png_opacity = 0.3)

# Make an elevation shading layer with dark valleys and light peaks (not essential but I like it!)
suppressMessages(elevation_overlay <- elevation_shade(DEM, elevation_palette = c("#000000", "#FFFFFF"), png_opacity = 0.6))

# Calculate the 'rayshader' scene (see 'rayshader' documentation)
elmat <- matrix(
  raster::extract(DEM, raster::extent(DEM), method = 'bilinear'),
  nrow = ncol(DEM),
  ncol = nrow(DEM)
)

scene <- elmat %>%
  sphere_shade(sunangle = sunangle, texture = "bw") %>%
  add_overlay(elevation_overlay) %>%
  add_overlay(stamen_overlay)

scene %>%
  plot_map()

## Render the 'rayshader' scene
#rayshader::plot_3d(
#  scene,
#  elmat,
#  zscale = zscale,
#  solid = FALSE,
#  shadow = TRUE,
#  shadowdepth = -100
#)
```

<p style="margin-top: -35px;"></p>

Candidates for Selectboard 2021

|                             |                             |
|-----------------------------|-----------------------------|
| Dan Fraser      &nbsp;&nbsp;|                 &nbsp;&nbsp;|
| John Hall       &nbsp;&nbsp;| Rachel Edens    &nbsp;&nbsp;|
| Ally Tufenkjian &nbsp;&nbsp;| Julia Dalphin   &nbsp;&nbsp;|

<div id="stream" style="text-align:center">
<a id="aud1_play" href="/vision#" onclick="(function(evt) { evt.target.innerHTML='Now Playing'; evt.target.onclick = null; window.userTriggered=true; }) (event || window.event)">Play</a>
&nbsp;<br /><br />
<!--a href="http://www.w3.org/2010/05/video/mediaevents.html" target="_blank">Media Help</a-->
<audio id="aud1" controls="true" preload="auto" muted="true" style="width:280px;">
<source src="/vision/vision.mp3" />
</audio>
</div>

<br /><br />

<em>Victorious Hartford</em> or <em>Victory For Hartford</em> means discovering and building strength from within our community. By strength, I mean creative problem solving, health, resilience, trust and well-being. We can see that the challenges of the recent past and of the near future will continue to push our current resources and systems to the absolute limit of their capacity to respond and to stabilize. We, and by we I mean the entire community, must consciously and continuously re-evaluate and re-invest in our collective capabilities ── enhancing our strengths, shining light on our blind spots and working through our shortcomings ── in order to overcome these challenges.

Hartford is a land of three rivers cradling countless flora and fauna. Trees cover a cumulative 23 square miles of the total 45 square mile expanse that is Hartford. Across that expanse, almost ten thousand souls live and labor and depend on the stability of the land beneath their feet and the social and economic systems that bind our realities together. These systems are as fragile as any forest ecosystem, possibly more so, and like any river, great care and maintenance and awareness are required to keep the water flowing to every place and every mouth that needs it.

As issues within our community come to the surface, be they challenges to our economic well-being, environmental health, social cohesion, or otherwise, we must honestly evaluate strategies to deepen our understanding of these challenges and plot a course of action that will fundamentally improve the ways that the Town relates to its personnel and to its residents. I understand that both the process of acknowledging real issues, and the genesis of effective strategies must be driven by the groups on which those issues have the greatest impact. We need to frequently remind ourselves of that dynamic value and accept that when we allow the powers of choice and action to stagnate, we lose the ability to adapt to disasters of all forms, be they biological, climactic, or economic. Every person in our community must be granted access and agency over these powers of governance in order for all of us to be victorious.
