---
title: "Events And Shade: Part 2"
author: John
date: '2020-11-19'
draft: false
categories:
  - GL
tags:
  - GL
  - Ray-Tracing
  - Shadows
  - SvelteJS
  - SvelteGL
  - Terrain
  - WebGL
---

<div id="simulation-app-2" role="main" style="width: 100%;">&nbsp;</div>

Events & Shade Across Time & Terrain: Part 2
============================================

```{r setup, cache = FALSE, echo = FALSE, include = FALSE}
options(warn = -1)
working_dir <- getwd()
print(working_dir)
if (!require("knitr")) {
    install.packages("knitr")
    library("knitr")
}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)

# Load reticulate package for running python
if (!require("png")) {
    install.packages("png")
}

library(png)

if (!require(akima) ||
    !require(reshape2) ||
    !require(viridis)) {
    install.packages(c("akima", "ggplot2", "reshape2", "viridis"))
}

if (!require(magrittr) || !require(rgdal) || !require(rayshader)) {
    install.packages(c("magrittr", "maptools", "rgdal", "rgeos", "sf", "sp"))
    devtools::install_github("tylermorganwall/rayimage@7a9a138e10e19119c88e960f9cfb191d1fdae002", update = "never")
    devtools::install_github("tylermorganwall/terrainmeshr@e112055e47033508cc45c8246b8dc0a0e94920f7", update = "never")
    devtools::install_github("tylermorganwall/rayshader@d0c9bd94be95c44eff6e7d8da5eadff070dc11db", update = "never")
}

library(magrittr)
library(akima)
library(ggplot2)
library(reshape2)
library(rayshader)
library(viridis)

DISPLACE_MULTIPLY <- 0.125
```


```{r images, cache = FALSE, echo = FALSE, results = "asis"}
if (file.exists("data/terrain-small.png")) {
    small_input <- "data/terrain-small.png"
} else {
    small_input <- file.choose()
}

image_test <- data.frame(png::readPNG(small_input))

if (file.exists("data/terrain-colormap.png")) {
    color_input <- "data/terrain-colormap.png"
} else {
    color_input <- file.choose()
}

image_color <- data.frame(png::readPNG(color_input))

if (file.exists("data/terrain-large.png")) {
    large_input <- "data/terrain-large.png"
} else {
    large_input <- file.choose()
}

image_gray <- data.frame(png::readPNG(large_input))

```


```{r map_vert_data, cache = FALSE, echo = FALSE}

mapHeight <- function(image_map, image_width, image_height) {
    # make a 1.0 x 1.0 plane starting at (0, 0, 0)... z is up
    heights <- matrix(1, nrow = image_height, ncol = image_width)

    for (y in 1:image_height) {
        for (x in 1:image_width) {
            heights[y, x] <- round(image_map[y, x], 6)
        }
    }

    heights
}
```


```{r preview_map_data, cache = FALSE, echo = FALSE, results = "asis"}
# import desert heightmap demo for test data
localtif <- raster::raster("data/dem_01.tif")

#And convert it to a matrix:
elmat <- raster_to_matrix(localtif)

heightmap <- mapHeight(image_test, length(image_test[1,]), length(image_test[, 1]))

#print(nrow(heightmap))   # width of the heightmap
#print(ncol(heightmap))   # height of the heightmap

# print without tabs so R markdown will add it to the DOM tree of the resulting page
cat(paste('<div id="data_in_html"><script type="application/json">', jsonlite::toJSON(heightmap), '\n</script>\n</div>', sep = ""))

# I'm going to substitute this heightmap matrix for the volcano matrix that was used in the example below,
# so I need to know use the volcano max as a multiplier against heightmap data (which is normalized)
heightmap_max <- max(apply(heightmap, c(1, 2), max))
#print(heightmap_max)
elmat_max <- max(apply(elmat, c(1, 2), max))
#print(elmat_max)
height_mult <- elmat_max / heightmap_max
#print(height_mult)

for (y in 1:nrow(heightmap)) {
    for (x in 1:ncol(heightmap)) {
        heightmap[y, x] <- heightmap[y, x] * (height_mult / 10)
    }
}

#print(max(apply(heightmap, c(1, 2), max)))
```


```{r heights, cache = TRUE, echo = FALSE}
# Map terrain heights from the highdensity image
mapTerrainCoords <- function(image_map, image_width, image_height) {
    # make a 2.0 x 2.0 plane, starting from at (-1, -1, 0) ... z is up
    heights <- matrix(1, nrow = image_height, ncol = image_width)
    height_array <- array(dim = c(image_height, image_width))
    vertices <- list()
    normals <- list()
    normal_array <- array(dim = c(image_height, image_width, 3))
    textureCoords <- list()

    # map heights
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            height <- round(image_map[y, x], 6)
            heights[y, x] <- height
        }
    }

    # map vetices, normals, texture
    i <- 0
    for (y in 1:image_height) {
        height_array_x <- c()
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            height_array_x <- c(height_array_x, heights[y, x])
            # vertices (x, y, z)
            # PNG y-coordinate system is reverse of WebGL
            vertices[[i]] <- c(2.0 * round((x / image_width - (1.0 / image_width) - 0.5), 6), 2.0 * round((y / image_height - (1.0 / image_height) - 0.5), 6), heights[(image_width + 1) - y, x] * DISPLACE_MULTIPLY)  # z is up
            # normals
            normals[[i]] <- c(0.0, 0.0, 1.0)
            # texture coordinates (s, t)
            textureCoords[[i]] <- c(round((x / image_width - ((image_width - x) / (image_width * image_width))), 6), round(1.0 - (y / image_height - ((image_height - y) / (image_height * image_height))), 6)) # go to flip the y
        }

        # start mapping z normals
        for (x in 1:image_width) {
            normal_array[y, x,] <- c(0.0, 0.0, round(height_array_x[x], 6))
        }

        height_array[y,] <- height_array_x
    }

    # Now re-calculate x and y normals based on neighboring heightmap coords
    i <- 0
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            # PNG y-coordinate system is reverse of WebGL
            if (x < image_width) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - y, x + 1, 3]
                xa <- (za - zb) # / image_width
                # print(paste0(za, " - ", zb,  " = ", xa))
                # normal_array[y, x,] <- c(round(xa, 6), 0.0, 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(round(xa, 6), normal_array[(image_width + 1) - y, x, 2], normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
            if (y < image_height) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - (y + 1), x, 3]
                ya <- (za - zb) # / image_height
                # print(paste0(za, " - ", zb,  " = ", ya))
                # normal_array[y, x,] <- c(0.0, round(ya, 6), 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(normal_array[(image_width + 1) - y, x, 1], round(ya, 6), normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
        }
    }

    list(
        heights = height_array,
        vertices = vertices,
        normals = normals,
        normal_array = normal_array,
        textureCoords = textureCoords
    )
}

#heights <- mapHeight(image_gray, length(image_gray[1,]), length(image_gray[, 1])) # start with the simple height mapper for now
map <- mapTerrainCoords(image_gray, nrow(image_gray), ncol(image_gray))

heights <- map$heights
vertices <- map$vertices
normals <- map$normals
textureCoords <- map$textureCoords
```


```{r normal_map, cache = FALSE, echo = FALSE, results = "asis"}

jsonlite::write_json(heights, "data/planar-terrain-heights.json")

if (file.exists("data")) {
  normal_test_file <- "data/planar-terrain-normals.png"
} else {
  normal_test_file <- file.choose(new = TRUE)
}

# test write to png
png::writePNG(map$normal_array, normal_test_file)

cat(paste0('<img id="heightmap" src="/posts/', normal_test_file, '" style="float: right; width:256px; padding: 1%;" />'))
```

In part 2, I continue to employ `mapTerrainCoords` that was developed in [Generating Planar Terrain Mesh](posts/terrain) to generate the terrain height maps and normal maps, as well as the ray-traced shading methods provided by <a href="https://github.com/tylermorganwall/rayshader" target="_blank">rayshader</a> that were explored in [Events And Shade: Part 1](posts/simulate-events-and-shade-1) to produce the shaded texture map.

```{r rayshaded_map, cache = FALSE, echo = TRUE, results = "asis"}
xparts <- 2
yparts <- 2
width <- ncol(heights)
height <- nrow(heights)
quad_width <- floor(width / xparts)
quad_height <- floor(height / yparts)

image_shaded <- array(dim = c(height, width, 3))

for (j in 1:yparts) {
    for (i in 1:xparts) {
        quad_width_start <- ((i - 1) * quad_width)
        quad_width_end <- (i * quad_width)
        quad_height_start <- ((j - 1) * quad_height)
        quad_height_end <- (j * quad_height)
        elmat <- matrix(1, nrow = quad_height, ncol = quad_width)

        for (y in 1:quad_height) {
            for (x in 1:quad_width ) {
                elmat[x, y] <- heights[quad_height_start + y, quad_width_start + x] * height_mult
            }
        }

        sunangle <- 45

        revx_elmat <- matrix(1, nrow = nrow(elmat), ncol = ncol(elmat))
        for (y in 1:nrow(elmat)) {
            for (x in 1:ncol(elmat)) {
                revx_elmat[y, x] <- elmat[(nrow(elmat) - y) + 1, x]
            }
        }

        terrain_shadow <- (revx_elmat %>%
            ray_shade(
                maxsearch = floor(sqrt(nrow(elmat)^2 + ncol(elmat)^2)) / 64,
                zscale = 100, sunaltitude = 45, sunangle = sunangle, lambert = TRUE,
                multicore = TRUE)
        )

        if (file.exists("data")) {
            terrain_shademap <- paste0("data/terrain-rayshaded-", (i - 1), "", (j - 1), ".png")
        } else {
            terrain_shademap <- file.choose(new = TRUE)
        }

        # image function and ggplot are slow; save with PNG function from PNG package.
        png::writePNG(t(terrain_shadow), terrain_shademap)

        for (y in 1:quad_height) {
            for (x in 1:quad_width ) {

                # combine shadow with colormap, with a value ratio of 1:2 respectively
                image_shaded[quad_height_start + y, quad_width_start + x, ] <- c(
                    ((terrain_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (0 * width)]),
                    ((terrain_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (1 * width)]),
                    ((terrain_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (2 * width)])
                )
            }
        }
    }
}

if (file.exists("data")) {
    rayshaded_file <- paste0("data/terrain-shaded-colormap.png")
} else {
    rayshaded_file <- file.choose(new = TRUE)
}

png::writePNG(image_shaded, rayshaded_file)

cat(paste0('<img id="shademap" src="/posts/', rayshaded_file, '" style="width: 100%;" />'))
```

