---
title: "Events And Shade: Part 1"
author: John
date: '2020-08-29'
draft: false
categories:
  - GL
tags:
  - GL
  - Ray-Tracing
  - Shadows
  - SvelteJS
  - SvelteGL
  - Terrain
  - WebGL
---

<div id="simulation-app" role="main" style="width: 100%;">
&nbsp;
</div>

Events & Shade Across Time & Terrain: Part 1
============================================

```{r setup, cache = FALSE, echo = FALSE, include = FALSE}
options(warn = -1)
working_dir <- getwd()
print(working_dir)
if (!require("knitr")) {
    install.packages("knitr")
    library("knitr")
}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)

# Load reticulate package for running python
if (!require("png")) {
    install.packages("png")
}

library(png)

if (!require(akima) ||
    !require(reshape2) ||
    !require(viridis)) {
    install.packages(c("akima", "ggplot2", "reshape2", "viridis"))
}

if (!require(magrittr) || !require(rgdal) || !require(rayshader)) {
    install.packages(c("magrittr", "maptools", "rgdal", "rgeos", "sf", "sp"))
    devtools::install_github("tylermorganwall/rayimage@7a9a138e10e19119c88e960f9cfb191d1fdae002", update = "never")
    devtools::install_github("tylermorganwall/terrainmeshr@e112055e47033508cc45c8246b8dc0a0e94920f7", update = "never")
    devtools::install_github("tylermorganwall/rayshader@d0c9bd94be95c44eff6e7d8da5eadff070dc11db", update = "never")
}

library(magrittr)
library(akima)
library(ggplot2)
library(reshape2)
library(rayshader)
library(viridis)

DISPLACE_MULTIPLY <- 0.125
```


```{r images, cache = FALSE, echo = FALSE, results = "asis"}
if (file.exists("data/terrain-small.png")) {
    small_input <- "data/terrain-small.png"
} else {
    small_input <- file.choose()
}

image_test <- data.frame(png::readPNG(small_input))

if (file.exists("data/terrain-colormap.png")) {
    color_input <- "data/terrain-colormap.png"
} else {
    color_input <- file.choose()
}

image_color <- data.frame(png::readPNG(color_input))

if (file.exists("data/terrain-large.png")) {
    large_input <- "data/terrain-large.png"
} else {
    large_input <- file.choose()
}

image_gray <- data.frame(png::readPNG(large_input))

cat(paste0('<img id="colormap" src="/posts/', color_input, '" style="float: right; width:256px; padding: 1%;" />'))
```

Given two texture images that describe the same terrain, one color and one grayscale, the first objective is to simulate the perceived time-of-day by altering the color texture with light and shade effects. The ray-tracing-based shade algorithm that I will use comes from the blog post [Throwing Shade at the Cartographer Illuminati](https://www.tylermw.com/throwing-shade/) by Tyler Morgan-Wall.
<br />

```{r map_vert_data, cache = FALSE, echo = FALSE}

mapHeight <- function(image_map, image_width, image_height) {
    # make a 1.0 x 1.0 plane starting at (0, 0, 0)... z is up
    heights <- matrix(1, nrow = image_height, ncol = image_width)

    for (y in 1:image_height) {
        for (x in 1:image_width) {
            heights[y, x] <- round(image_map[y, x], 6)
        }
    }

    heights
}
```

```{r preview_map_data, cache = FALSE, echo = FALSE, results = "asis"}
# import desert heightmap demo for test data
localtif <- raster::raster("data/dem_01.tif")

#And convert it to a matrix:
elmat <- raster_to_matrix(localtif)

heightmap <- mapHeight(image_test, length(image_test[1,]), length(image_test[, 1]))

#print(nrow(heightmap))   # width of the heightmap
#print(ncol(heightmap))   # height of the heightmap

# print without tabs so R markdown will add it to the DOM tree of the resulting page
cat(paste('<div id="data_in_html"><script type="application/json">', jsonlite::toJSON(heightmap), '\n</script>\n</div>', sep = ""))

# I'm going to substitute this heightmap matrix for the volcano matrix that was used in the example below,
# so I need to know use the volcano max as a multiplier against heightmap data (which is normalized)
heightmap_max <- max(apply(heightmap, c(1, 2), max))
#print(heightmap_max)
elmat_max <- max(apply(elmat, c(1, 2), max))
#print(elmat_max)
height_mult <- elmat_max / heightmap_max
#print(height_mult)

for (y in 1:nrow(heightmap)) {
    for (x in 1:ncol(heightmap)) {
        heightmap[y, x] <- heightmap[y, x] * (height_mult / 10)
    }
}

#print(max(apply(heightmap, c(1, 2), max)))
```


```{r throwing_shade, cache = TRUE, echo = TRUE, results= "asis"}
shadow <- matrix(1, nrow = nrow(heightmap), ncol = ncol(heightmap))
tanangheight <- matrix(1, nrow = nrow(heightmap), ncol = ncol(heightmap))

sunangle <- 45 / 180 * pi
angle <- -90 / 180 * pi
diffangle <- 90 / 180 * pi
numberangles <- 25
anglebreaks <- seq(angle, diffangle, length.out = numberangles)
maxdistance <- floor(sqrt(nrow(heightmap)^2 + ncol(heightmap)^2))

for (y in 1:nrow(heightmap)) {
    for (x in 1:ncol(heightmap)) {
        for (anglei in anglebreaks) {
            for (k in 1:maxdistance) {
                xcoord <- (x + cos(sunangle) * k)
                ycoord <- (y + sin(sunangle) * k)
                if (xcoord > nrow(heightmap) ||
                    ycoord > ncol(heightmap) ||
                    xcoord < 0 ||
                    ycoord < 0) {
                    break
                } else {
                    tanangheight[y, x] <- heightmap[y, x] + tan(anglei) * k
                    if (all(
                        c(heightmap[ceiling(ycoord), ceiling(xcoord)],
                          heightmap[ceiling(ycoord), floor(xcoord)],
                          heightmap[floor(ycoord), ceiling(xcoord)],
                          heightmap[floor(ycoord), floor(xcoord)]
                        ) < tanangheight[y, x]
                    )) next
                    if (tanangheight[y, x] < bilinear(1:ncol(heightmap), 1:nrow(heightmap), heightmap, x0 = xcoord, y0 = ycoord)$z) {
                        shadow[y, x] <- shadow[y, x] - 1 / length(anglebreaks)
                        break
                    }
                }
            }
        }
    }
}
```


```{r shadow_image, cache = TRUE, echo = FALSE, results = "asis"}
shadowmap <- matrix(1, nrow = nrow(shadow), ncol = ncol(shadow))

for (y in 1:nrow(shadow)) {
    for (x in 1:ncol(shadow)) {
        shadowmap[y, x] <- shadow[y, x]
    }
}

if (file.exists("data")) {
    shadow_image <- "data/terrain-shadow.png"
} else {
    shadow_image <- file.choose(new = TRUE)
}

# test write to png
png::writePNG(shadowmap, shadow_image)

cat(paste0('<img src="/posts/', shadow_image, '" style="float: right; width:256px; padding: 1%;" />'))
```

The R implementation of this particular method is pretty slow, as explained in the post about the algorithm referenced above, even on the small 128x128 example that was used. So, the author created a package called [`rayshader`](https://www.rayshader.com/) that imports a C implementation of the same. Here's how it works:

```{r rayshader_package, cache = TRUE, echo = TRUE}
elmat <- matrix(1, nrow = nrow(image_test), ncol = ncol(image_test))

for (y in 1:nrow(image_test)) {
    for (x in 1:ncol(image_test)) {
        elmat[x, y] <- heightmap[y, x]
    }
}

sunangle <- 45

elmat %>%
    ray_shade(
        maxsearch = floor(sqrt(nrow(elmat)^2 + ncol(elmat)^2)) / 64,
        zscale = 100, sunaltitude = 45, sunangle = sunangle, lambert = TRUE,
        multicore = TRUE) %>%
    plot_map()
```

Following the examples is pretty straight forward and incredibly useful once you understand the parameters. However, the actual data I want to render and superimpose on the colormap is quite large, so I will break it down into quadrants before processing:

```{r heights, cache = TRUE, echo = FALSE}
# Map terrain heights from the highdensity image
mapTerrainCoords <- function(image_map, image_width, image_height) {
    # make a 2.0 x 2.0 plane, starting from at (-1, -1, 0) ... z is up
    heights <- matrix(1, nrow = image_height, ncol = image_width)
    height_array <- array(dim = c(image_height, image_width))
    vertices <- list()
    normals <- list()
    normal_array <- array(dim = c(image_height, image_width, 3))
    textureCoords <- list()

    # map heights
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            height <- round(image_map[y, x], 6)
            heights[y, x] <- height
        }
    }

    # map vetices, normals, texture
    i <- 0
    for (y in 1:image_height) {
        height_array_x <- c()
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            height_array_x <- c(height_array_x, heights[y, x])
            # vertices (x, y, z)
            # PNG y-coordinate system is reverse of WebGL
            vertices[[i]] <- c(2.0 * round((x / image_width - (1.0 / image_width) - 0.5), 6), 2.0 * round((y / image_height - (1.0 / image_height) - 0.5), 6), heights[(image_width + 1) - y, x] * DISPLACE_MULTIPLY)  # z is up
            # normals
            normals[[i]] <- c(0.0, 0.0, 1.0)
            # texture coordinates (s, t)
            textureCoords[[i]] <- c(round((x / image_width - ((image_width - x) / (image_width * image_width))), 6), round(1.0 - (y / image_height - ((image_height - y) / (image_height * image_height))), 6)) # go to flip the y
        }

        # start mapping z normals
        for (x in 1:image_width) {
            normal_array[y, x,] <- c(0.0, 0.0, round(height_array_x[x], 6))
        }

        height_array[y,] <- height_array_x
    }

    # Now re-calculate x and y normals based on neighboring heightmap coords
    i <- 0
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            # PNG y-coordinate system is reverse of WebGL
            if (x < image_width) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - y, x + 1, 3]
                xa <- (za - zb) # / image_width
                # print(paste0(za, " - ", zb,  " = ", xa))
                # normal_array[y, x,] <- c(round(xa, 6), 0.0, 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(round(xa, 6), normal_array[(image_width + 1) - y, x, 2], normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
            if (y < image_height) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - (y + 1), x, 3]
                ya <- (za - zb) # / image_height
                # print(paste0(za, " - ", zb,  " = ", ya))
                # normal_array[y, x,] <- c(0.0, round(ya, 6), 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(normal_array[(image_width + 1) - y, x, 1], round(ya, 6), normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
        }
    }

    list(
        heights = height_array,
        vertices = vertices,
        normals = normals,
        normal_array = normal_array,
        textureCoords = textureCoords
    )
}

#heights <- mapHeight(image_gray, length(image_gray[1,]), length(image_gray[, 1])) # start with the simple height mapper for now
map <- mapTerrainCoords(image_gray, nrow(image_gray), ncol(image_gray))

heights <- map$heights
vertices <- map$vertices
normals <- map$normals
textureCoords <- map$textureCoords
```


```{r rayshader_high_dim, cache = TRUE, echo = TRUE, results = "asis"}
xparts <- 2
yparts <- 2
width <- ncol(heights)
height <- nrow(heights)
quad_width <- floor(width / xparts)
quad_height <- floor(height / yparts)

image_shaded <- array(dim = c(height, width, 3))

#print(paste0("image width: ", ncol(image_shaded)))
#print(paste0("image height: ", nrow(image_shaded)))

for (j in 1:yparts) {
    for (i in 1:xparts) {
        quad_width_start <- ((i - 1) * quad_width)
        quad_width_end <- (i * quad_width)
        quad_height_start <- ((j - 1) * quad_height)
        quad_height_end <- (j * quad_height)
        elmat <- matrix(1, nrow = quad_height, ncol = quad_width)

        for (y in 1:quad_height) {
            for (x in 1:quad_width ) {
                elmat[x, y] <- heights[quad_height_start + y, quad_width_start + x] * height_mult
            }
        }

        sunangle <- 45

        revx_elmat <- matrix(1, nrow = nrow(elmat), ncol = ncol(elmat))
        for (y in 1:nrow(elmat)) {
            for (x in 1:ncol(elmat)) {
                revx_elmat[y, x] <- elmat[(nrow(elmat) - y) + 1, x]
            }
        }

        single_angle_shadow <- (revx_elmat %>%
            ray_shade(
                maxsearch = floor(sqrt(nrow(elmat)^2 + ncol(elmat)^2)) / 64,
                zscale = 100, sunaltitude = 45, sunangle = sunangle, lambert = TRUE,
                multicore = TRUE)
        )

        if (file.exists("data")) {
            rayshader_test_file <- paste0("data/terrain-rayshaded-", (i - 1), "", (j - 1), ".png")
        } else {
            rayshader_test_file <- file.choose(new = TRUE)
        }

        # image function and ggplot are slow; save with PNG function from PNG package.
        png::writePNG(t(single_angle_shadow), rayshader_test_file)

        cat(paste0('<img id="shademap', (i - 1), "", (j - 1), '" src="/posts/', rayshader_test_file, '" style="width: 100%" />'))

        for (y in 1:quad_height) {
            for (x in 1:quad_width ) {
                # combine shadow with colormap, with a value ratio of 1:2 respectively
                image_shaded[quad_height_start + y, quad_width_start + x, ] <- c(
                    ((single_angle_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (0 * width)]),
                    ((single_angle_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (1 * width)]),
                    ((single_angle_shadow[x, y] / 2 + 0.5) * image_color[quad_height_start + y, quad_width_start + x + (2 * width)])
                )
            }
        }
    }
}

if (file.exists("data")) {
    rayshaded_file <- paste0("data/terrain-shaded-colormap.png")
} else {
    rayshaded_file <- file.choose(new = TRUE)
}

## test write to png
png::writePNG(image_shaded, rayshaded_file)

cat(paste0('<img id="shademap" src="/posts/', rayshaded_file, '" style="width: 100%" />'))
```

I have made use of the `mapTerrainCoords` method that was developed in [Generating Planar Terrain Mesh](posts/terrain) to generate the terrain heights and normals, and then write them out to a PNG and JSON file, respectively:

```{r idx_data, cache = FALSE, echo = FALSE, results = "asis"}

jsonlite::write_json(heights, "data/planar-terrain-heights.json")

if (file.exists("data")) {
    normal_test_file <- "data/planar-terrain-normals.png"
} else {
    normal_test_file <- file.choose(new = TRUE)
}

# test write to png
png::writePNG(map$normal_array, normal_test_file)

cat(paste0('<img id="heightmap" src="/posts/', normal_test_file, '" style="width: 100%" />'))
```
