---
title: "Generating Planar Terrain Mesh"
author: John
date: '2020-07-31'
draft: false
categories:
  - GL
tags:
  - GL
  - Shader
  - SvelteJS
  - SvelteGL
  - WebGL
---

<div id="terrain-app" role="main" style="width: 100%;">
&nbsp;
</div>

## Generating Planar Terrain Mesh

```{r setup, cache = FALSE, echo = FALSE, include=FALSE}
options(warn = -1)
working_dir <- getwd()
print(working_dir)
if (!require("knitr")) {
    install.packages("knitr")
    library("knitr")
}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)

# Load reticulate package for running python
if (!require("png")) {
    install.packages("png")
}
library(png)

DISPLACE_MULTIPLY <- 0.125
```

Given a gray scale heightmap, I need a planar triangulated mesh to serve as the base for a terrain object + shader to be used in the above webapp.

I already have some tools to abstract a heightmap from an image. I'm going to start with a smaller version of the terrain image file which I've converted to grayscale:

```{r image, cache = FALSE, echo = TRUE, results = "asis"}
if (file.exists("data/terrain-small.png")) {
    small_input <- "data/terrain-small.png"
} else {
    small_input <- file.choose()
}

image <- data.frame(png::readPNG(small_input))

cat(paste0('<img src="/posts/', small_input, '" style="float: right; width:256px; padding: 1%;" />'))
```

<div class="r-data">

```{r preview_1, cache = FALSE, echo = FALSE}
length(image[,1])   # width of the map
length(image[1,])   # height of the map
str(image)
```
</div>

Now for every point in the image, I need to create coordinates for a 3d vertex (x, y, z), a normal for each vertex and texture coordinates (s, t) for each vertex:

```{r map_vert_data, cache = FALSE, echo = TRUE}

mapHeight <- function (image_map, image_width, image_height) {
    # make a 1.0 x 1.0 plane starting at (0, 0, 0)... z is up
    heights <- matrix(nrow = image_width, ncol = image_height)

    for (y in 1:image_height) {
        for (x in 1:image_width) {
            heights[y, x] <- round(image_map[y, x], 6)
        }
    }

    for (y in 1:image_height) {
        for (x in 1:image_width) {
            # vertices (x, y, z)
            # z is "up", and png y coordinate system is reverse of WebGL
            print(paste0(round((x / image_width), 6), ", ", round(y / image_height, 6), ", ", heights[(image_width + 1) - y, x]))
            # normals
            print(paste0(0.0, ", ", 0.0, ", ", 1.0)) # z is "up"
            # texture coordinates (s, t)
            print(paste0(round((x / image_width), 6), ", ", round(y / image_height, 6)))
        }
    }

    heights
}
```

<div class="r-data">

```{r preview_map_data, cache = FALSE, echo = FALSE}
heightmap <- mapHeight(image, length(image[1,]), length(image[, 1]))

length(heightmap[1,])   # width of the heightmap
length(heightmap[,1])   # height of the heightmap
```
</div>


The following line will convert the heightmap frame into JavaScript or JSON data and optionally print the results within HTML tags:
```{r data_in_html, echo = TRUE, results = "asis"}
require(jsonlite)

jsonHeightmap <- toJSON(heightmap)

# print without tabs so R markdown will add
# it to the DOM tree of the resulting page
cat(paste0(
    '<div id="data_in_html">\n',
    '<script type="application/json">\n', jsonHeightmap, '\n', '</script>\n',
    '</div>'
))

```

```{r image_test, cache = FALSE, echo = FALSE, results = "asis"}

if (file.exists("data")) {
    vertex_test_file <- "data/terrain-heights.png"
} else {
    vertex_test_file <- file.choose(new = TRUE)
}

# test write to png
png::writePNG(heightmap, vertex_test_file)

cat(paste0('<img src="/posts/', vertex_test_file, '" style="float: right; width:256px; padding: 1%;" />'))

```

The input and output test looks good, so time to work on a larger (medium sized) terrain image.
```{r image_2, cache = FALSE, echo = TRUE, results = "asis"}
if (file.exists("data/terrain-medium.png")) {
    medium_input <- "data/terrain-medium.png"
} else {
    medium_input <- file.choose()
}

image <- data.frame(png::readPNG(medium_input))
```

<div class="r-data">

```{r preview_2, cache = FALSE, echo = TRUE}
length(image[1,])   # width of the map
length(image[,1])   # height of the map
str(image)
```
</div>

Once the vertices, the normals, and the texture coordinates have been built for this new map, I will need indices to tie everything together. The sequence of the indices will trace the vertices into triangular faces:

```{r idx_data, cache = TRUE, echo = TRUE}

mapTerrainCoords <- function (image_map, image_width, image_height) {
    # make a 2.0 x 2.0 plane, starting from at (-1, -1, 0) ... z is up
    heights <- matrix(nrow = image_width, ncol = image_height)
    height_array <- array(dim = c(image_height, image_width))
    vertices <- list()
    normals <- list()
    normal_array <- array(dim = c(image_height, image_width, 3))
    textureCoords <- list()

    # map heights
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            heights[y, x] <- round(image_map[y, x], 6)
        }
    }

    # map vetices, normals, texture
    i <- 0
    for (y in 1:image_height) {
        height_array_x <- c()
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            height_array_x <- c(height_array_x , heights[y, x])
            # vertices (x, y, z)
            # png y-coordinate system is reverse of WebGL
            vertices[[i]] <- c(2.0 * round((x / image_width - (1.0 / image_width) - 0.5), 6), 2.0 * round((y / image_height - (1.0 / image_height) - 0.5), 6), heights[(image_width + 1) - y, x] * DISPLACE_MULTIPLY)  # z is up
            # normals
            normals[[i]] <- c(0.0, 0.0, 1.0)
            # texture coordinates (s, t)
            textureCoords[[i]] <- c(round((x / image_width - ((image_width - x) /  (image_width * image_width))), 6), round(1.0 - (y / image_height - ((image_height - y) /  (image_height * image_height))), 6)) # go to flip the y
        }

        # start mapping z normals
        for (x in 1:image_width) {
            normal_array[y, x,] <- c(0.0, 0.0, round(height_array_x[x], 6))
        }

        height_array[y,] <- height_array_x
    }

    total_points <- i

    print(total_points)

    print("heights:")
    print(length(height_array[1,]) * length(height_array[,1]))
    print(length(height_array[1,]))

    print("vertices:")
    print(length(vertices))
    print(length(vertices[[1]]))

    print("textureCoords:")
    print(length(textureCoords))
    print(length(textureCoords[[1]]))

    # Now re-calculate x and y normals based on neighboring heightmap coords
    i <- 0
    for (y in 1:image_height) {
        for (x in 1:image_width) {
            i <- i + 1 # increment before to start on 1
            # png y-coordinate system is reverse of WebGL
            if (x < image_width) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - y, x + 1, 3]
                xa <- (za - zb) # / image_width
                # print(paste0(za, " - ", zb,  " = ", xa))
                # normal_array[y, x,] <- c(round(xa, 6), 0.0, 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(round(xa, 6), normal_array[(image_width + 1) - y, x, 2], normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
            if (y < image_height) {
                za <- normal_array[(image_width + 1) - y, x, 3]
                zb <- normal_array[(image_width + 1) - (y + 1), x, 3]
                ya <- (za - zb) # / image_height
                # print(paste0(za, " - ", zb,  " = ", ya))
                # normal_array[y, x,] <- c(0.0, round(ya, 6), 0.0)
                normal_array[(image_width + 1) - y, x,] <- c(normal_array[(image_width + 1) - y, x, 1], round(ya, 6), normal_array[(image_width + 1) - y, x, 3])
                normals[[i]] <- normal_array[(image_width + 1) - y, x,]
            }
        }
    }

    print("normals:")
    print(length(normals))
    print(length(normals[[1]]))
    print(length(normal_array[,,1]))    # total sets of normal coords
    print(length(normal_array[,1,1]))   # normal coords per image row
    print(length(normal_array[1,1,]))   # normal coordinate dimensions
    list(
        heights = height_array,
        vertices = vertices,
        normals = normals,
        normal_array = normal_array,
        textureCoords = textureCoords
    )
}

image_width <- length(image[1,])
image_height <- length(image[, 1])
map <- mapTerrainCoords(image, image_width, image_height)
```

```{r normal_preview, cache = TRUE, echo = FALSE, results = "asis"}
heights <- map$heights
vertices <- map$vertices
normals <- map$normals
textureCoords <- map$textureCoords

if (file.exists("data")) {
   normal_test_file <- "data/terrain-normals.png"
} else {
    normal_test_file <- file.choose(new = TRUE)
}

# test write to png
png::writePNG(map$normal_array, normal_test_file)

cat(paste0('<img id="heightmap" src="/posts/', normal_test_file, '" style="float: right; width:256px; padding: 1%;" />'))
```

Now I have some idea of what this is supposed to look like, so I can verify whether my indexing algorithm works...
```{r idx_data_2, cache = TRUE, echo = TRUE}
currentRow <- 1
indices <- c()

indices <- c(indices, (image_width - 2))
indices <- c(indices, (image_width - 1 + image_width))
indices <- c(indices, (image_width - 1 + (image_width - 1)))

indices <- c(indices, (image_width - 1 + image_width))
indices <- c(indices, (image_width - 2))
indices <- c(indices, (image_width - 1))

print(indices)
```

It's looking like this might be recursive, so I will need a helper function...
<pre><code class="r idx_knit_recurssion_fail">
knitTriangleIndices <- function(img_width, img_height) {
    ind <- c()
    width <- img_width - 1
    row <- 1

    ...

    knitTriangleIndices(..., ...)

    ...
}
</code></pre>

Hours later, I decide my force powers are not strong enough for recursion in R, hence the following for-loop.
```{r idx_knit_for_loop, cache = TRUE, echo = TRUE}

knitTriangleIndices <- function(img_width, img_height) {
    indices <- c()
    width <- img_width - 1
    row <- 1

    for (r in 1:(img_height - 1)) {
        offset <- (img_width * (row - 1))
        while (width > 0) {
            idx1 <- (width - 1) + offset
            idx2 <- (width + img_width) + offset
            idx3 <- (width + (img_width - 1)) + offset

            idx4 <- (width + img_width) + offset
            idx5 <- (width - 1) + offset
            idx6 <- (width) + offset

            indices <- c(indices, idx1, idx2, idx3, idx4, idx5, idx6)

            width <- (width - 1)
        }
        width <- img_width - 1
        row <- row + 1
    }

    indices
}

```

Finally, all of this data must be saved to json and then stitched together into a JS module which defines a new `terrain` (heightmap) object that can be imported by [`@sveltejs/gl`](https://github.com/Real-Currents/SvelteGL/tree/real/currents), with the form:
<pre><code>
    new Geometry({
        position: {
            data: new Float32Array([ /* array of vertices ... */ ]
                .flat(Infinity)),
            size: 3
        },

        normal: {
            data: new Float32Array([ /* array of normals ... */ ]
                .flat(Infinity)),
            size: 3
        },

        uv: {
            data: new Float32Array([ /* array of textureCoords ... */ ]
                .flat(Infinity)),
            size: 2
        }
    }, {
        index: new Uint32Array([ /* array of indices returned by knitTriangleIndices */ ]
        )
    });

</code></pre>

Now this turns out to be a [ginormous MJS file](/posts/data/terrain.mjs) (~ 20MB), so I will ultimately use the smaller heightmap as the basis of `terrain` object imported by `@sveltejs/gl`. However, I will use the higher density height data exported by the normalmap (the blue PNG above) as well as create a json file for reference and use it to place the markers on the terrain later:

```{r json_write, cache = TRUE}
jsonlite::write_json(heights, "data/planar-terrain-heights.json")
jsonlite::write_json(vertices, "data/planar-terrain-verts.json")
jsonlite::write_json(normals, "data/planar-terrain-norms.json")
jsonlite::write_json(textureCoords, "data/planar-terrain-uvs.json")
jsonlite::write_json(knitTriangleIndices(image_width, image_height), "data/planar-terrain-idx.json")
```
