---
title: "Visualizing Terrain in SveltR"
author: John
date: '2020-04-07'
draft: false
categories:
  - GL
tags:
  - GIS
  - GL
  - Landscape
  - SvelteJS
  - SvelteGL
  - Terrain
  - WebGL
---

<div id="terrain-app" role="main" style="width: 100%; min-width: 800px; margin-left: -86px;">
&nbsp;
</div>

## Visualizing Terrain

```{r setup, cache = FALSE, echo = FALSE, results = "asis"}
options(warn = -1)
working_dir <- getwd()
#print(working_dir)
if (!require("knitr")) {
    install.packages("knitr")
    library("knitr")
}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)

# Load reticulate package for running python
if (!require("png")) {
    install.packages("png")
}

library(png)

if (file.exists("data/heightmap-small.png")) {
  input <- "data/heightmap-small.png"
} else {
  input <- file.choose()
}

image <- data.frame(png::readPNG(input))

cat(paste0('<img id="heightmap" src="/posts/', input, '" style="float: right; width:256px; padding: 1%;" />'))
```

As I mentioned before if the output target is HTML, R has some methods for turning R data (the same height map data as used in the [Visualizing R Data](posts/visualize-r) post)...

```{r load_data, echo = FALSE}
heightmap <- data.frame(c())
if (print(file.exists(input))) {
  heightmap <- data.frame(png::readPNG(input))
  is.num <- sapply(heightmap, is.numeric)     # for all numeric values...
  heightmap[is.num] <- lapply(heightmap[is.num], round, 6)  # ... lower precision to 6 decimals
}
```

<div class="r-data">

```{r display_data, echo = FALSE}
head(heightmap)
```

... into JavaScript or JSON data and printing the results within HTML tags:
```{r data_in_html, echo = FALSE}

cat(paste(
'<div id="data_in_html">\n',
'   <script type="application/json">\n\t\t',
jsonlite::toJSON(heightmap), '\n',
'   </script>\n',
'</div>',
sep=""))

```
</div>

Which is great for direct data analysis, but if we intend to use the image as a height map for terrain visualization, we can skip the data ingestion. As mentioned in the post about [GLSL shaders](posts/glsl), I've extended the GL extension to SvelteJS ([`@sveltejs/gl`](https://github.com/sveltejs/gl/pull/28)), so that a set of custom vertex and fragment shaders can use a normal map (given as the same PNG image above) to displace the vertices of a planar polygon mesh in order to create a terrain-like model.

`TerrainApp.svelte`:

    <script>
        import { onMount } from 'svelte';
        import * as GL from '@sveltejs/gl';
        import terrainFrag from './shaders/custom/terrain-frag.glsl';
        import terrainVert from './shaders/custom/terrain-and-light-vert.glsl';

        export let title;

        export let color = '#F7C77B';

        let webgl;
        let normalmap;
        let process_extra_shader_components = (gl, material, model) => {
            console.log("Process Extra Shader Components");
            const program = material.program;

            if (material.vertName === "terrain-vert") {
                // console.log(material.vertName);

                const heigthAdjMultLocation = gl.getUniformLocation(program, "height_adjustment");
                gl.uniform1f(heigthAdjMultLocation, 0.5);

                // uniform vec3 light_direction; // normalized direction in eye
                const lightDirectionLocation = gl.getUniformLocation(program, "light_direction");
                gl.uniform3fv(lightDirectionLocation, new Float32Array([light.x,light.y,light.z]));
                // uniform vec3 light_ambient_color;
                const ambientLightLocation = gl.getUniformLocation(program, "light_ambient_color");
                gl.uniform3fv(ambientLightLocation, new Float32Array([...normalizeColor(light.color)]));
                // uniform vec3 light_diffuse_color;
                const diffuseLightLocation = gl.getUniformLocation(program, "light_diffuse_color");
                gl.uniform3fv(diffuseLightLocation, new Float32Array([...normalizeColor(light.color)]));
                // uniform vec3 light_specular_color;
                const specularLightLocation = gl.getUniformLocation(program, "light_specular_color");
                gl.uniform3fv(specularLightLocation, new Float32Array([...normalizeColor(light.color)]));

                // uniform float material_specular_exponent;
                const specularExpLocation = gl.getUniformLocation(program, "material_specular_exponent");
                gl.uniform1f(specularExpLocation, 0.99);

            }

        };

        onMount(() => {
            // ...

            normalmap = new GL.Texture(document.getElementById("heightmap").src, { width: 512, height: 512 });

        });
    </script>


    <GL.Scene bind:gl={webgl} backgroundOpacity=1.0 process_extra_shader_components={process_extra_shader_components}>

        ...

        <!-- ground -->
        <GL.Mesh
                geometry={GL.terrain()}
                location={[0, -h/2, 0]}
                rotation={[-rot, 0, 0]}
                scale={h}
                frag={terrainFrag}
                vert={terrainVert}
                uniforms={{ color: adjustColor(color, h), alpha: 1.0, normalmap: normalmap }}
        />

    </GL.Scene>


Here's the two shaders that have been applied to the above scene...

`terrain-and-light-vert.glsl`:

    #define NAME terrain-vert

    #define C_ZERO 0.0
    #define C_QUARTER 0.25
    #define C_HALF 0.5
    #define C_ONE 1.0

    #define DISPLACE_MULTIPLY 0.5

    // texture containing elevation data
    //uniform sampler2D heightMap;
    //uniform sampler2D bumpmap;
    uniform sampler2D normalmap;

    uniform float height_adjustment;

    in vec3 position;

    in vec3 normal;

    // available when texture maps are used
    // (bumpmap, colormap, normalmap...)
    in vec2 uv;

    out vec3 v_normal;

    out vec2 v_textureCoords;

    uniform vec3 light_direction; // normalized direction in eye
    uniform vec3 light_ambient_color;
    uniform vec3 light_diffuse_color;
    uniform vec3 light_specular_color;
    uniform float material_specular_exponent;

    out vec3 v_directional_light_shading;

    vec3 directional_light_shading (vec3 normal) {
        vec3 computed_shade = vec3(C_ZERO, C_ZERO, C_ZERO);
        vec3 halfplane_vector = vec3(C_ZERO, C_ONE, C_ZERO);
        vec3 nlight_direction = normalize(light_direction); // normalized direction in eye
        vec3 nlight_halfplane = normalize(nlight_direction + halfplane_vector); // normalized half-plane vector
        float ndotL; // dot product of normal & light direction
        float ndotH; // dot product of nomral and & half-plane vector

        ndotL = max(C_ZERO, dot(normal, nlight_direction));
        computed_shade += light_ambient_color * vec3(C_HALF, C_HALF, C_HALF);
        computed_shade += ndotL * light_diffuse_color * vec3(C_ONE, C_ONE, C_ONE);

        // The resolution of the vertex shader is not fine enough
        // for specular lighting, but this is the equation
        ndotH = max(C_ZERO, dot(normal, nlight_halfplane));
        if (ndotH > C_ZERO) {
            computed_shade += pow(ndotH, material_specular_exponent) * vec3(C_ONE, C_ONE, C_ONE) * light_specular_color;
        }

        return computed_shade;
    }

    void main() {
        vec3 displacement = texture(normalmap, uv).rgb;

        vec3 displace_along_verticle = normal * displacement * (DISPLACE_MULTIPLY * height_adjustment);

        vec3 displaced_position = position + displace_along_verticle;

        vec3 displace_along_normal = vec3(normal * displacement);

        v_normal = DISPLACE_MULTIPLY * normal;

        #if defined(has_normalmap)
        v_normal = DISPLACE_MULTIPLY * displace_along_normal;
        #endif

        v_textureCoords = uv;

        v_directional_light_shading = directional_light_shading((MODEL_INVERSE_TRANSPOSE * vec4(v_normal, C_ZERO)).xyz);

        gl_Position = PROJECTION * VIEW * MODEL * vec4(displaced_position, C_ONE);
    }


`terrain-frag.glsl`:

    /* Uniforms supplied by Svelte materials: */
    uniform vec3 color;
    uniform vec3 emissive;
    uniform float alpha;
    uniform float specularity;
    uniform sampler2D bumpmap;
    uniform sampler2D colormap;
    uniform sampler2D emissivemap;
    uniform sampler2D specularitymap;
    uniform vec3 FOG_COLOR;
    uniform float FOG_DENSITY;

    #define NAME terrain-frag

    #define C_ZERO 0.0
    #define C_QUARTER 0.25
    #define C_HALF 0.5
    #define C_ONE 1.0

    in vec3 v_directional_light_shading;

    in vec3 v_normal;

    in vec2 v_textureCoords;

    /* Varying for this shader: */
    out mediump vec4 fragColor;

    void main () {
    //    fragColor = vec4(color, 1.0);
        fragColor = vec4(v_directional_light_shading * color, C_ONE);

        if (alpha == C_ZERO || (C_ZERO <= alpha && alpha <= C_ONE)) {
            fragColor.a *= alpha;
        }
    }

